```{r, echo = FALSE}
w0 <- 9
```

## Week 9 lecture 1

### Challenge I

1. Using the midpoint rule with $N = 5$ nodes per variable, approximate
\[
I = \int_1^3 \ldots \int_1^3 \prod_{i = 1}^4 \left\{1 + (x_i - 2)^2\right\} \text{d}x_1 \ldots \text{d}x_4
\]
and estimate its relative absolute error.

<details><summary>**Solution**</summary>
```{r, eval = FALSE, echo = week > w0}
N <- 5
d <- 4
b <- 3
a <- 1
h <- (b - a) / N
x <- a + h * (1:N - .5)
xx <- lapply(1:d, function(i) x)
X <- expand.grid(xx)
w <- h^d
f <- apply(1 + (X - 2)^2, 1, prod)
midpoint <- w * sum(f)
true <- (8/3)^d
rel_err <- abs((true - midpoint) / true)
rel_err
```
</details>

## Week 9 lecture 2

### Challenge I

1. Consider $Y \sim \text{Exp}(\lambda)$, so that its pdf is 
\[
f(y \mid \lambda) = \lambda \text{exp}(-\lambda y) \quad \text{for}~y > 0,
\]
where $\lambda$ is a random variable with pdf
\[
f(\lambda) = 4 \lambda \text{exp}(-2 \lambda) \quad \text{for}~\lambda > 0.
\]
Use Laplace's method to find $f(y)$, where
\[
f(y) = \int_0^\infty f(y \mid \lambda) f(\lambda) \text{d} \lambda.
\]

<details><summary>**Solution**</summary>

- We first notice that, given Equation (4.4), we take $f()$ to be $-\log f(y, \lambda)$ and $x$ to be $\lambda$, as that's what we're integrating out.

- Then we get $f(y, \lambda) = f(y \mid \lambda) f(\lambda) = \lambda \text{e}^{-\lambda y} \times 4 \lambda \text{e}^{-2 \lambda} = 4 \lambda^2 \text{e}^{-\lambda(2 + y)}$

- So $-\log f(y, \lambda) = -2 \log \lambda + \lambda(2 + y) - \log 4$

- Then $-\dfrac{\partial \log f(y, \lambda)}{\partial \lambda} = -\dfrac{2}{\lambda} + (2 + y)$, which is minimised at $\tilde \lambda = 2 / (2 + y)$

- Also $-\dfrac{\partial^2 \log f(y, \lambda)}{\partial \lambda^2} = 2 / \lambda^2$ and $2 / \tilde \lambda^2 = 2 / (4 / (2 + y)^2) = (2 + y)^2 / 2$

- So, if $g''(\lambda)$ denotes $-\partial^2 \log f(y, \lambda) / \partial \lambda^2$, then
\begin{align*}
I_n &= f(y, \tilde \lambda) \sqrt{\dfrac{2 \pi}{g''(\tilde \lambda)}}\\
&= 4 \hat \lambda^2 \text{e}^{-\hat \lambda(2 + y)} \sqrt{2\pi \times 2(2 + y)^{-2}} = \dfrac{32 \sqrt{\pi}}{\text{e}^2(2 + y)^3}
\end{align*}

</details>

### Challenges II

1. Use Monte Carlo integration with $N = 10^3$ Monte Carlo samples to approximate
\[
\int_{-1}^1 x^2 \text{d} x
\]
and then calculate the relative absolute error of your approximation.

<details><summary>**Solution**</summary>
```{r, eval = FALSE, echo = week > w0}
N <- 1e3
V <- 2
mc <- V * mean(runif(N, -1, 1)^2)
true <- 2/3
```
</details>

### Challenges III

1. Approximate
\[
I = \int_1^3 \ldots \int_1^3 \prod_{i = 1}^4 \left\{1 + (x_i - 2)^2\right\} \text{d}x_1 \ldots \text{d}x_4
\]
using Monte Carlo sampling and a Monte Carlo sample size of $N = 10^4$ and estimate the relative absolute error of your approximation.

<details><summary>**Solution**</summary>
```{r, eval = FALSE, echo = week > w0}
N <- 10^4
d <- 4
x <- matrix(runif(N * d, 1, 3), N)
V <- 2^d
I_hat <- V * mean(apply(1 + (x - 2)^2, 1, prod))
I_hat2 <- V * mean(exp(rowSums(log1p((x - 2)^2))))
true <- (8/3)^d
rel_err <- abs((true - I_hat) / true)
rel_err
```
</details>

# Chapter 5

<!-- ## Week 9 lecture 3 -->

<!-- ### Challenges I -->

<!-- ```{r, echo = FALSE} -->
<!-- n <- 10 -->
<!-- set.seed(202203171) -->
<!-- y <- sort(round(runif(n), 2)) -->
<!-- ``` -->

<!-- 1. Consider a random variable $Y$ with pdf -->
<!-- \[ -->
<!-- f(y \mid \theta) = \dfrac{\Gamma(2\theta) [y(1 - y)]^{\theta - 1}}{[\Gamma(\theta)]^2} \quad 0 \leq y \leq 1 -->
<!-- \] -->
<!-- and for $\theta > 0$. The log-likelihood for an independent sample, $y_1, \ldots, y_n$, is then given by -->
<!-- \[ -->
<!-- \log f(\mathbf{y} \mid \theta) = n\big[\log\big(\Gamma(2\theta)\big) - 2\log\big(\Gamma(\theta)\big)\big] + (\theta - 1) \sum_{i = 1}^n \left\{\log y_i + \log(1 - y_i)\right\}. -->
<!-- \] -->
<!-- Write a function, `negloglik(theta, y)` to evaluate the *negative* of the log-likelihood above, and then for the data -->
<!-- \[ -->
<!-- `r paste(y, collapse = ', ')` -->
<!-- \] -->
<!-- find the maximum likelihood estimate $\hat \theta$ using `optimize()`, assuming that $1 \leq \hat \theta \leq 3$. [Recall that `lgamma(x)` evaluates $\log\big(\Gamma(x)\big)$ in `R` if $x =$ `x`.] -->

<!-- <details><summary>**Solution**</summary> -->
<!-- ```{r, eval = FALSE, echo = week > w0} -->
<!-- negloglik <- function(theta, y) { -->
<!--   # Function to evaluate negative log-likelihood -->
<!--   # theta is a scalar -->
<!--   # y is a vector -->
<!--   # returns a scalar -->
<!--   n <- length(y) -->
<!--   loglik <- n * (lgamma(2 * theta) - 2 * lgamma(theta)) -->
<!--   loglik <- loglik + (theta - 1) * sum(log(y) + log(1 - y)) -->
<!--   -loglik -->
<!-- } -->

<!-- y <- c(.16, .35, .37, .46, .67, .69, .79, .8, .91, .92) -->
<!-- negloglik(2, y) -->
<!-- optimize(negloglik, c(1, 3), y = y) -->
<!-- ``` -->
<!-- </details> -->
