```{r, echo = FALSE}
w0 <- 10
```

## Week 10 lecture 1

### Challenges I

1. Consider modelling the sample $y_1, \ldots, y_n$ independent as Exp($\lambda$) random variables. Given that their log likelihood is
\[
\log f(\mathbf{y} \mid \lambda) = n \log \lambda - \lambda \sum_{i = 1}^n y_i
\]
and then that
\[
\dfrac{\partial \log f(\mathbf{y} \mid \lambda)}{\partial \lambda} = \dfrac{n}{\lambda} - \sum_{i = 1}^n y_i \quad \text{and} \quad \dfrac{\partial^2 \log f(\mathbf{y} \mid \lambda)}{\partial \lambda^2} = -\dfrac{n}{\lambda^2}
\]
write functions called `nd1(lambda, y)` and `nd2(lambda, y)` that evaluate the first and second derivatives of the *negative* log-likelihood, i.e. of $-\log f(\mathbf{y} \mid \lambda)$.

<details><summary>**Solution**</summary>
```{r, eval = FALSE, echo = week > w0}
nd1 <- function(lambda, y) {
  # Function to evaluate first derivative of exponential
  # negative log-likelihood w.r.t. lambda
  # lambda is a scalar
  # y is a vector
  # returns a scalar
  n <- length(y)
  sum(y) - n / lambda
}

nd2 <- function(lambda, y) {
  # Function to evaluate second derivative of exponential
  # negative log-likelihood w.r.t. lambda
  # lambda is a scalar
  # y is a vector
  # returns a scalar
  n <- length(y)
  n / lambda^2
}
```
</details>

2. Find the first Newton step given a starting value of $\lambda_0 = 1$ and the sample
\[
0.2, 0.5, 0.6, 1.0, 1.8, 2.5
\]

<details><summary>**Solution**</summary>
```{r, eval = FALSE, echo = week > w0}
y <- c(0.2, 0.5, 0.6, 1.0, 1.8, 2.5)
-nd1(1, y) / nd2(1, y)
```
</details>

### Challenges II

```{r, echo = FALSE}

nd1 <- function(lambda, y) {
  # Function to evaluate first derivative of exponential
  # negative log-likelihood w.r.t. lambda
  # lambda is a scalar
  # y is a vector
  # returns a scalar
  n <- length(y)
  sum(y) - n / lambda
}

nd2 <- function(lambda, y) {
  # Function to evaluate second derivative of exponential
  # negative log-likelihood w.r.t. lambda
  # lambda is a scalar
  # y is a vector
  # returns a scalar
  n <- length(y)
  matrix(n / lambda^2, 1, 1)
}
```

1. Consider modelling the sample $y_1, \ldots, y_n$ independent as Exp($\lambda$) random variables. Given that their log likelihood is
\[
\log f(\mathbf{y} \mid \lambda) = n \log \lambda - \lambda \sum_{i = 1}^n y_i,
\]
and that $\lambda > 0$, write a function called `nd0(lambda, y)` that evaluates the *negative* log-likelihood, i.e. $-\log f(\mathbf{y} \mid \lambda)$.

<details><summary>**Solution**</summary>
```{r, eval = FALSE, echo = week > w0}
nd0 <- function(lambda, y) {
  # Function to evaluate exponential negative log-likelihood
  # lambda is a scalar
  # y is a vector
  # returns a scalar
  n <- length(y)
  -n * log(lambda) + lambda * sum(y)
}
```
</details>

2. Use `nlminb()` with `nd0()`, `nd1()` and `nd2()` to find $\hat \lambda$, the maximum likelihood estimate, given independent sample
\[
0.2, 0.5, 0.6, 1.0, 1.8, 2.5
\]
and a starting value of $\lambda_0 = 1$. [Note that you will need to ensure that `nd2()` returns a matrix, albeit a trivial $1 \times 1$ one. Using `as.matrix()` is probably easiest.]

<details><summary>**Solution**</summary>
```{r, eval = FALSE, echo = week > w0}
nd1 <- function(lambda, y) {
  # Function to evaluate first derivative of exponential
  # negative log-likelihood w.r.t. lambda
  # lambda is a scalar
  # y is a vector
  # returns a scalar
  n <- length(y)
  sum(y) - n / lambda
}

nd2 <- function(lambda, y) {
  # Function to evaluate second derivative of exponential
  # negative log-likelihood w.r.t. lambda
  # lambda is a scalar
  # y is a vector
  # returns a scalar
  n <- length(y)
  matrix(n / lambda^2, 1, 1)
}

y <- c(.2, .5, .6, 1, 1.8, 2.5)
lambda_0 <- 1
nlminb(lambda_0, nd0, nd1, nd2, y = y)
```
</details>

<!-- 3. Confirm your result with `optimize()` and `optim(..., method = 'Brent)`. -->

<!-- <details><summary>**Solution**</summary> -->
<!-- ```{r, eval = FALSE, echo = week > w0} -->
<!-- optimize(nd0, c(.5, 1.5), y = y) -->
<!-- optim(1, nd0, lower = .5, upper = 1.5, method = 'Brent', y = y) -->
<!-- ``` -->
<!-- </details> -->

3. Experiment with different starting values $\lambda_0$ in order to investigate the sensitivity of Newton's method to starting values (on this occasion)

<details><summary>**Solution**</summary>
```{r, eval = FALSE, echo = week > w0}
lambda0_seq <- 10^seq(-4, 4)
sapply(lambda0_seq, function(x) nlminb(x, nd0, nd1, nd2, y = y)$par)
```
</details>

## Week 10 lecture 2

### Challenges I

1. Consider modelling the sample $y_1, \ldots, y_n$ independent as Exp($\text{e}^\tau$) random variables. Given that their log likelihood is
\[
\log f(\mathbf{y} \mid \tau) = n \tau - \text{e}^\tau \sum_{i = 1}^n y_i,
\]
use `nlminb()` by supplying the negative log-likelihood and its first and second derivatives w.r.t. $\tau$ to find $\hat \tau$, its maximum likelihood estimate, given independent sample
\[
0.2, 0.5, 0.6, 1.0, 1.8, 2.5
\]
and a starting value of $\tau_0 = 1$.

<details><summary>**Solution**</summary>
```{r, eval = FALSE, echo = week > w0}
nd0_tau <- function(tau, y) {
  # function to evaluate Exp(exp(tau)) neg. log lik.
  # tau is a scalar
  # y is a vector
  # returns a scalar
  n <- length(y)
  - n * tau + exp(tau) * sum(y)
}

y0 <- c(.2, .5, .6, 1, 1.8, 2.5)
tau0 <- 0
nd0_tau(tau0, y0)

nd1_tau <- function(tau, y) {
  # function to 1st deriv. w.r.t. to tau
  # of Exp(exp(tau)) neg. log lik.
  # tau is a scalar
  # y is a vector
  # returns a scalar
  n <- length(y)
  - n + exp(tau) * sum(y)
}

nd2_tau <- function(tau, y) {
  # function to 2nd deriv. w.r.t. to tau
  # of Exp(exp(tau)) neg. log lik.
  # tau is a scalar
  # y is a vector
  # returns a scalar
  as.matrix(exp(tau) * sum(y))
}

nd1_tau(tau0, y0)
nd2_tau(tau0, y0)

fit <- nlminb(tau0, nd0_tau, nd1_tau, nd2_tau, y = y0)
tau_hat <- fit$par
exp(tau_hat)
```
</details>

2. Confirm the result with `optimize()`.

<details><summary>**Solution**</summary>
```{r, eval = FALSE, echo = week > w0}
optimize(nd0_tau, c(-1, 1), y = y0)
```
</details>

## Week 10 lecture 3

### Challenges I

```{r, echo = FALSE}
n <- 10
y <- c(1.8, 2.1, 2.3, 2.4, 2.5, 2.7, 3.1, 4.3, 4.3, 4.4)
```

1. Consider the sample $y_1, \ldots, y_n$ given by
\[
`r paste(y, collapse = ', ')`
\]
as $n = `r n`$ independent realisations from the Normal($\mu$, $\sigma^2$) with $\mu$ unknown and $\sigma^2 = 1$, i.e. with pdf
\[
f(y \mid \mu, \sigma) = \frac{1}{\sqrt{2 \pi \sigma^2}} \text{e}^{-(y - \mu)^2/2\sigma^2}.
\]
Suppose that $\sigma^2 = 1$. Find $\log f(\mathbf{y} \mid \mu = 1, \sigma^2 = 1) = \sum_{i = 1}^n \log f(y_i \mid \mu = 1, \sigma^2 = 1)$.

<details><summary>**Solution**</summary>
```{r, eval = FALSE, echo = week > w0}
y <- c(1.8, 2.1, 2.3, 2.4, 2.5, 2.7, 3.1, 4.3, 4.3, 4.4)
sum(dnorm(y, 1, 1, log = TRUE))
```
</details>

2. Write a function in `R`, `negloglik(mu, y)`, that evaluates $-\log f(\mathbf{y} \mid \mu, \sigma^2 = 1)$ for `mu` $= \mu$ and `y` $=y$. 

<details><summary>**Solution**</summary>
```{r, eval = FALSE, echo = week > w0}
negloglik <- function(mu, y) {
 # Function to evaluate derivative negative log-likelihood
  # mu is a scalar
  # y is a vector
  # returns a scalar
  n <- length(y)
  n * log(2 * pi) / 2 + sum((y - mu)^2) / 2
}

negloglik(1, y)
```
</details>

3. Find $-\partial \log f(\mathbf{y} \mid \mu, \sigma^2 = 1) / \partial \mu$ and write a function in `R`, `negloglik_d1(mu, y)`, that evaluates $-\partial \log f(\mathbf{y} \mid \mu, \sigma^2 = 1) / \partial \mu$  for `mu` $= \mu$ and `y` $=y$.

<details><summary>**Solution**</summary>
```{r, eval = FALSE, echo = week > w0}
negloglik_d1 <- function(mu, y) {
  # Function to evaluate derivative of 
  # negative log-likelihood w.r.t. mu
  # mu is a scalar
  # y is a vector
  # returns a scalar
  -sum(y - mu)
}

negloglik_d1(1, y)
```
</details>

4. Use `optim()` in `R` to find $\hat \mu$, the maximum likelihood estimate of $\mu$, using the BFGS method. You should supply `negloglik_d1(mu, y)` to evaluate the first derivative w.r.t. $\mu$, and use $\mu_0 = 1$ as a starting value.

<details><summary>**Solution**</summary>
```{r, eval = FALSE, echo = week > w0}
fit_bfgs <- optim(1, negloglik, negloglik_d1, y = y, method = 'BFGS')
mu_hat <- fit_bfgs$par
```
</details>

5. Evaluate $-\partial \log f(\mathbf{y} \mid \mu, \sigma^2 = 1) / \partial \mu$ at $\hat \mu$ to verify that $\hat \mu$ is the m.l.e.

<details><summary>**Solution**</summary>
```{r, eval = FALSE, echo = week > w0}
negloglik_d1(mu_hat, y)
mean(y)
```
</details>

<!-- ## Week 11 lecture 1 -->

<!-- ### Challenges II -->

<!-- ```{r, echo = FALSE} -->
<!-- set.seed(202203241) -->
<!-- n <- 10 -->
<!-- y <- c(1.8, 2.1, 2.3, 2.4, 2.5, 2.7, 3.1, 4.3, 4.3, 4.4) -->
<!-- ``` -->

<!-- Consider the sample $y_1, \ldots, y_n$ given by -->
<!-- \[ -->
<!-- `r paste(y, collapse = ', ')` -->
<!-- \] -->
<!-- as $n = `r n`$ independent realisations from the Normal($\mu$, $\sigma^2$) with $\mu$ and $\sigma^2$ unknown, i.e. with pdf -->
<!-- \[ -->
<!-- f(y \mid \mu, \sigma) = \frac{1}{\sqrt{2 \pi \sigma^2}} \text{e}^{-(y - \mu)^2/2\sigma^2}. -->
<!-- \] -->

<!-- 1. Write a function in `R`, `negloglik(pars, y)`, that evaluates $-\log f(\mathbf{y} \mid \mu, \sigma^2)$ for `pars[1]` $= \mu$, `pars[2]` $= \sigma$  and `y` $=y$. -->

<!-- <details><summary>**Solution**</summary> -->
<!-- ```{r, eval = FALSE, echo = week > w0} -->
<!-- negloglik <- function(pars, y) { -->
<!--  # Function to evaluate derivative negative log-likelihood -->
<!--   # pars is a 2-vector, where -->
<!--   # pars[1] is mu and -->
<!--   # pars[2] is sigma -->
<!--   # y is a vector -->
<!--   # returns a scalar -->
<!--   mu <- pars[1] -->
<!--   sigma <- pars[2] -->
<!--   if (sigma <= 0) -->
<!--     return(1e6) -->
<!--   n <- length(y) -->
<!--   n * log(2 * pi) / 2 + n * log(sigma) + sum((y - mu)^2) / 2 / sigma^2 -->
<!-- } -->

<!-- y <- c(1.8, 2.1, 2.3, 2.4, 2.5, 2.7, 3.1, 4.3, 4.3, 4.4) -->

<!-- negloglik(c(1.5, 1.5), y) -->

<!-- -sum(dnorm(y, 1.5, 1.5, log = TRUE)) -->
<!-- ``` -->
<!-- </details> -->

<!-- 2. Find $-\nabla \log f(\mathbf{y} \mid \mu, \sigma)$, i.e. the 2-vector comprising $-\partial \log f(\mathbf{y} \mid \mu, \sigma) / \partial \mu$ and $-\partial \log f(\mathbf{y} \mid \mu, \sigma) / \partial \sigma$, and then write a function in `R` that evaluates $-\nabla \log f(\mathbf{y} \mid \mu, \sigma)$. -->

<!-- <details><summary>**Solution**</summary> -->
<!-- ```{r, eval = FALSE, echo = week > w0} -->
<!-- negloglik_d1 <- function(pars, y) { -->
<!--   # Function to evaluate derivative of  -->
<!--   # negative log-likelihood w.r.t. mu and sigma -->
<!--   # pars is a 2-vector, where -->
<!--   # pars[1] is mu and -->
<!--   # pars[2] is sigma -->
<!--   # y is a vector -->
<!--   # returns a 2-vector -->
<!--   mu <- pars[1] -->
<!--   sigma <- pars[2] -->
<!--   n <- length(y) -->
<!--   out <- numeric(2) -->
<!--   res <- y - mu -->
<!--   out[1] <- -sum(res) / sigma^2 -->
<!--   out[2] <- n / sigma - sum(res^2) / sigma^3 -->
<!--   out -->
<!-- } -->

<!-- negloglik_d1(c(1.5, 1.5), y) -->
<!-- ``` -->
<!-- </details> -->

<!-- 3. Use the BFGS method with `optim()` in `R` to find $\hat \mu$ and $\hat \sigma$, the maximum likelihood estimates of $\mu$ and $\sigma$. Use $\mu_0 = 1.5$ and $\sigma_0 = 1.5$ as starting values. -->

<!-- <details><summary>**Solution**</summary> -->
<!-- ```{r, eval = FALSE, echo = week > w0} -->
<!-- fit_bfgs <- optim(c(1.5, 1.5), negloglik, negloglik_d1, y = y, method = 'BFGS') -->
<!-- mles <- fit_bfgs$par -->
<!-- ``` -->
<!-- </details> -->

<!-- 4. Use the $-\nabla \log f(\mathbf{y} \mid \mu, \sigma)$ evaluated at $\hat \mu$ and $\hat \sigma$ to verify that the m.l.e.s have been found. -->

<!-- <details><summary>**Solution**</summary> -->
<!-- ```{r, eval = FALSE, echo = week > w0} -->
<!-- negloglik_d1(mles, y) -->
<!-- mean(y) -->
<!-- sqrt(var(y) * (length(y) - 1) / length(y)) -->
<!-- ``` -->
<!-- </details> -->

<!-- ## Week 10 lecture 3 -->

<!-- ### Challenges I -->

<!-- 1. Consider the alternative parameterisation of the Weibull pdf in terms of parameters $(\log \lambda, \log k)$ given by  -->
<!-- \[ -->
<!-- f(y \mid \tilde \lambda, \tilde k) =  -->
<!-- \frac{\exp(\tilde k)}{\exp(\tilde \lambda)}\left(\frac{y}{\exp(\tilde \lambda)}\right)^{\exp(\tilde k)-1}e^{-(y/\exp(\tilde \lambda))^{\exp(\tilde k)}} \quad y > 0, -->
<!-- \] -->
<!-- where $-\infty < \tilde \lambda, \tilde k < \infty$. Write a function in `R`, `weib2_d0(pars, y, mult = 1)`, that evaluates the scaled log-likelihood of the model, i.e. $m \log f(\mathbf{y} \mid \tilde \lambda, \tilde k)$, where `pars` is a 2-vector such that `pars[1]` $= \tilde \lambda$ and `pars[2]` $= \tilde k$, `y` is the data $\mathbf{y}$ and `mult` $= m$ is a multiplier as for `weib_d0()`. -->

<!-- <details><summary>**Solution**</summary> -->
<!-- ```{r, eval = FALSE, echo = week > w0} -->
<!-- weib2_d0 <- function(pars, y, mult = 1) { -->
<!--   # Function to evaluate re-parameterised Weibull log-likelihood -->
<!--   # pars is a vector -->
<!--   # pars[1] is $\tilde \lambda$ -->
<!--   # pars[2] is $\tilde k$ -->
<!--   # y can be scalar or vector -->
<!--   # mult is a scalar defaulting to 1; so -1 returns neg. log likelihood -->
<!--   # returns a scalar -->
<!--   n <- length(y) -->
<!--   epars <- exp(pars) -->
<!--   out <- n * (pars[2] - epars[2] * pars[1]) +  -->
<!--     (epars[2] - 1) * sum(log(y)) - sum((y / epars[1])^epars[2]) -->
<!--   mult * out -->
<!-- } -->
<!-- ``` -->
<!-- </details> -->

<!-- 2. Find the maximum likelihood estimates of $(\tilde \lambda, \tilde k)$ using the Nelder-Mead method, starting at $(\tilde \lambda_0, \tilde k_0) = \big(\log(1.6), \log(0.6)\big)$. -->

<!-- <details><summary>**Solution**</summary> -->
<!-- ```{r, eval = FALSE, echo = week > w0} -->
<!-- fit_nm2 <- optim(log(c(1.6, 0.6)), weib2_d0, y = y0, mult = -1) -->
<!-- fit_nm2$par -->
<!-- ``` -->
<!-- </details> -->

<!-- 3. Confirm that these are related to those of the original parameterisation through the logarithm. -->

<!-- <details><summary>**Solution**</summary> -->
<!-- ```{r, eval = FALSE, echo = week > w0} -->
<!-- fit_nm$par -->
<!-- exp(fit_nm2$par) -->
<!-- ``` -->
<!-- </details> -->
